\textbf{ A Sphere Model for Atrial Fibrillation}

### 2017-02-21 Literature Review/MSCI POSTER draft.tex
poster stuff

\documentclass[twocolumn, a1paper]{article}
 %\usepackage[scale=0.78,size=a1]{beamerposter}
%\setlength{\paperheight}{33.1in}
%\setlength{\paperwidth}{23.4in}
\usepackage{cite}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{commath}
\usepackage{adjustbox}
%\usepackage[sc]{mathpazo}
\usepackage[margin=0.5in]{geometry}
%\usepackage{multicol}
\usepackage{mathtools}
\usepackage{abstract}
\usepackage[compact]{titlesec}
\DeclareMathOperator{\sech}{sech}
\newcommand{\matr}[1]{\mathbf{#1}}
%\usepackage{placeins}
%\usepackage{textcomp}
\graphicspath{ {H:/} }

@@

Each heart beat originates as an electric signal in the sinoatrial (SA) node that propagates first into the atria, then through the atrioventricular (AV) node, through Purkinje fibres and finally from the ventricular endocardium to the ventricular epicardium. The electric signal is conducted in the cardiac muscle cells thanks to the polarisation mechanism of the cell membrane. 
When a muscle cell (myocyte) is at rest, there is a built up potential difference between the outside and the inside of the cell (polarisation).

@@

AF is usually produced  by an calcified muscle cells (fibrosis) in the cardiac tissue which generates a process called reentry. During reentry the electric signal wavefront propagation breaks. It has been shown a clear link between fibrosis and AF
Even though many treatments have been developed, the mechanisms behind this condition are not completely understood and AF remains a major topic in medical research. The current research revolves around three possible approaches:

\smartdiagram[bubble diagram]{Current research,
 Physiological \\ models, Image-Based \\ Models, \textbf{ CA Models}, antiarrhythmic \\ drugs) t}

@@

"Simple Model for Identifying Critical Regions in Atrial Fibrillation"~\cite{Christensen}produced a simple 2D model for AF. The cardiac cells were arranged in a squared grid with horizontal connections, with a percentage of transversal connections (to reproduce the effect of fibrosis) and a percentage of unexcitable cells.

Electric impulse propagation rules:

%insert fucking rules

The results from this model show that there is a threshold value of vertical connections beyond which AF is produced spontaneously (without changing refractory period or conduction speed). This matches real data according to which AF is correlated with fibrosis. 
The model also shows that burning the area affected by AF effectively stops the arrhythmia, which has been recently discovered to be the case .

\section{\textbf{Implementing Restitution in Kim Christensen Model}} %METHODS

Mechanisms thought to be responsible for active cardiac arrhythmias are either enhanced/abnormal pulse formation or re-entry. For the former, simulations have included modelling the ionic currents of action potentials such as the Fenton-Karma model \cite{Grandi} \cite{Fenton} and modelling the impulse of an action potential using generic reaction-diffusion equations, such as the FitzHugh-Nagumo Model \cite{FitzHugh}. Most of these models are based on the pioneering work of Hodgkin and Huxley who described the action potential of the giant squid axon, which bears similarities to cardiac cells \cite{Hodgkin}. %CHECKKKKKKKKKKKKKKKKKKK
 For all of these models bidomain or monodomain approaches can be used to model the electrical propagation in myocardial tissue. % intracellularly????
  The bidomain model takes into account the anisotropy of intracellular and extracellular spaces. The monodomain model is a reduction of the bidomain model by assuming anisotropy ratios of intracellular and extracellular domains are equal---this is the model most commonly used in simulations today.%ET CETERA?
% ,   is the Many types of models have been used to simulate fibrillatory behaviour in the heart, from, and signals for contraction, to cellular automata and

Re-entry---the mechanism by which a propagating impulse fails to die out after normal activation of the heart---has been studied in depth via various circus reentry formulations: the Ring Model, the Leading circle model, the Figure-of-Eight model and the Spiral Wave model \cite{Antze} \cite{Tusscher}. The Ring model models a wave of propagation that moves along one, unidirectional pathway of heart tissue and returns to the origin of excitation, thus simulating a central area of inexcitable tissue. This simple model implies that the conduction velocity and the refractory period are both factors: the length of the circuit (pathlength) must be greater or equal to the wavelength (the distance travelled by an impulse during the refractory period). The smaller the wavelength, the greater the number of reentry circuits that can be accommodated during the refractory period to promote fibrillatory behaviour. See figure 2(c).

The Spiral wave model---which consists of a single rotating wavefront around a core, simulating tachycardia---has been observed in systems that have sites that have excited, refractory and resting states in both homogeneous and heterogeneous media \cite{Greenburg} \cite{Bub}. This is due to anisotropies in conduction velocity in the cardiac tissue. See figure 2(b) and figure 3. % for comparison to a normal wavefront. 
 Experimental demonstration of spiral waves has been achieved by in research by Davidenko et al and the concept is now seen to be an underlying driver of fibrillation; the paradigm appears to more closely correspond with many clinical and experimental observations than other models \cite{Davidenko} \cite{Comtois}. In one such observation, the presence of rotors was seen in 96\% of patients with AF, in a 2012 study \cite{afstudy}. The curvature of the spiral wave is related to the speed of the wave, the more curvature the slower the speed of propagation. High curvature, and far-field instability, leads to spiral wave breakage into daughter wavelets. This agrees with
%popular classical modes of thought regarding the initiation of fibrillatory contraction, multiple circuit reentry, which in this case is due to spiral breakup, and 
recent opinions where single, small reentry circuits or local generators have been found to promote fibrillatory activity---the core of the spiral \cite{Skanes} \cite{Nattel2}. The tip of the spiral wave has been seen to drift around the core, thus moving the whole spiral wave. Such meandering can be a factor in promoting fibrillatory activity \cite{Tusscher}.
% Recent observations are in contradiction with the classical viewpoint,  . This is also concurrent 
 Extending the spiral wave model to 3D gives rise to scroll waves which propagate through multiple layers of cells and give rise to more complex wave dynamics and is therefore hard to implement computationally.  %It has been seen that the if a spiral wave is in homogenous media then it will always return to its resting state .

\begin{figure}
\caption[short title]{Diagram showing a spiral wave reentry circuit. (a), a normal wavefront in the heart. (b) spiral wave, which simulates tachycardia, as can be seen by multiple propagating wavefronts in a small area. (c), turbulent activity due to break up of spiral wave giving rise to fibrillation \cite{Alonso}}
\centering
\includegraphics[width = 0.25\textwidth]{spiralbreak2tachy}
\end{figure}

@@

Recent research has shown that rotating wavefronts can be initiated in regions where the proportion of transverse coupling between strands of tissue is reduced to a critical value, leading to spontaneous fibrillatory activity, in contrast to normal 'linear' wavefronts. This can be seen in the rectilinear cellular automata model developed by Christensen et al and reflects current opinion: the fibrosis of cells, which reduces the coupling of electrical signals to other cells, leads to AF\cite{Christensen}. This research also supports targeted ablation to stop AF. Other studies suggest similar mechanisms such as the coupling of fibroblasts to atrial myocytes heterogeneously gives rise to spiral wave breakups \cite{Ozawa}.

@@

More cellular automata models have been researched by Siregar et al, where a 3D model of the heart, with simplified morphology is simulated and, in addition, a 2D CA model is compared qualitative model \cite{Siregar} \cite{Siregar2}. A hybrid I/O cellular automata approach has also been studied in research by Bartocci et al to describe cardiac tissue and improve the efficiency of the simulation process by parallelization \cite{Bartocci}.

 The impact of left and right atrial tissue electrical heterogeneities to promote AF has been studied in a biophysical model and it was seen that AF is more persistent in the right atria \cite{Luca}. Further development could be achieved by simulation from a cellular automata perspective. Modelling scroll waves and replicating the morphology of the heart with its heterogeneities more closely seem to be viable directions for further research. The effects of antiarrhythmic drugs on the heart during AF can be studied, along with the associated mechanism by which proarrhythmia occurs, and methods for its rectification. Developing ways to find CFAEs with novel electrogram methods via simulation could help targeted ablation, stopping AF more effectively and efficiently.

% Further developments could be obtained through this parallelization combined with 3D modelling, more accurately depicting the shape of the heart, making the code more efficient.

 %Many of these studies suggest targeted ablation can hinder or fully stop AF, therefore, electrogram-guided ablation to find CFAEs is a promising method it is concurrent with clinical observations. %electrogram-guided ablation to find CFAE.

%Most of these models only work one or two dimensionally, assuming that the heart is a 2D surface and not a 3D

%Cellular automaton models

%MAIN PAPER FROM KIM CHRISTIENSEN
%BIPOLAR ELECTROGRAM??
%METHODS TO POTENTIALLY STOP ATRIAL FIBRILLATION

%FUUUUUUUUUUUUURTHER METHODZZZZZZZZZZZZZZZZZZZZZZZZZZ

%\section{\textbf{\underline{Implications and Discussion}}}  %What our research project will be able to achieve, potentially

>>>>

\documentclass[twocolumn, a1paper, 20pt]{article}
 %\usepackage[scale=0.78,size=a1]{beamerposter}
%\setlength{\paperheight}{33.1in}
%\setlength{\paperwidth}{23.4in}
\usepackage{cite}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{commath}
\usepackage{adjustbox}
%\usepackage[sc]{mathpazo}
\usepackage[margin=0.5in]{geometry}
%\usepackage{multicol}
\usepackage{mathtools}
\usepackage{abstract}
\usepackage[compact]{titlesec}
\DeclareMathOperator{\sech}{sech}
\newcommand{\matr}[1]{\mathbf{#1}}
%\usepackage{placeins}
%\usepackage{textcomp}
\graphicspath{ {H:/} }

@@

Each heart beat originates as an electric signal in the sinoatrial (SA) node that propagates first into the atria, then through the atrioventricular (AV) node, through Purkinje fibres and finally from the ventricular endocardium to the ventricular epicardium. The electric signal is conducted in the cardiac muscle cells thanks to the polarisation mechanism of the cell membrane.

@@

When the electric impulse propagates a single muscle cell goes through three stages:

\begin{itemize}
  \item \textit{Excitable state}: the muscle cell is at rest with a negative built up potential.

  \item \textit{Excited state}: the muscle cell is excited by its excited neighbours and the Voltage of the cell is at its peak. 
  \item \textit{Refractory state} :the muscle cell goes through a phase during which it can't be excited again/
\end{itemize}

Where the refractory period depends on the last time a cell was excited.\\

AF is correlated to the amount of \textbf{ fibrosis} in the cardiac tissue which generates a process called reentry. \\
During reentry the electric signal wavefront propagation breaks. It has been shown a clear link between fibrosis and AF
Even though many treatments have been developed, the mechanisms behind this condition are not completely understood and AF remains a major topic in medical research. \\
The current research revolves around three possible approaches:

\smartdiagram[bubble diagram]{Current research,
 Physiological \\ models, Image-Based \\ Models, \textbf{ CA Models}, antiarrhythmic \\ drugs) }

@@

\subsection{Description of the model}

In the publication "a Simple Model for Identifying Critical Regions in Atrial Fibrillation"~\cite{Christensen}, Prof. Christensen produced a simple 2D model for AF. The cardiac cells were arranged in a squared grid with horizontal connections, with a percentage of \textbf{ transversal connections} (to reproduce the effect of fibrosis) and a percentage of \textbf{dysfunctional cells}.
Dysfunctional cells are excited by neighbours with a fixed probability.\\

\textbf{\textit{Propagation Rules the model}}\\

\begin{itemize}
  \item \textit{Excitable state}: at time step t=0, each cell at rest is an element in the grid

  \item \textit{Excited state}: at t=1, the cell is excited by one of its \textit{ 4 neighbour cells}, and it can excite one the excitable neighbour cells at the next time-step.
  \item \textit{Refractory state} :from t=2 onwards, the cell goes through the refractory state for t=T time-steps before becoming excitable again/
\end{itemize}

\subsection{Results}

The result show that AF naturally arises from the model itself.
\textbf{ It is important to note that this model is the only model ever created where arrhythmia naturally occurs from the structure of the heart} instead of being put in by hand in some way (as for instance adding by hand a new source of electric impulses.

 % insert picture here of the model with wavefronf broken etc

 In addition the simulations produced by the model show a correlation between fibrillation and fibrosis .
There is a threshold value of vertical connections beyond which AF is produced spontaneously (without changing refractory period or conduction speed). This matches real data according to which AF is correlated with fibrosis. 
The model also shows that burning the area affected by AF effectively stops the arrhythmia, which has been recently discovered to be the case.

%insert picture

\section{\textbf{Implementing Restitution in Kim Christensen Model}}

@@

@@
